Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-12-22 analysis/paper/London_HauloutBeahvior.Rmd
Andy VD comments incorporated

Ice-associated seals rely on sea ice for a variety of activities, including
breeding, molting, pupping, and resting. In the Arctic, many of these activities
occur in spring (April-June) as sea ice begins to melt and retreat northward.
Rapid acceleration of climate change in Arctic ecosystems is therefore of
concern as the quantity and quality of suitable habitat is forecast to decrease.
In this paper, we use data collected from satellite-linked bio-loggers deployed
between 2005 and 2020 to investigate the seasonal timing and environmental
factors affecting haul-out behavior by ice-associated seals (specifically, bearded,
ribbon, and spotted seals) in the Bering and Chukchi seas. In addition to
providing baseline data on phenology, these data also allow us to quantify
'availability', which is needed to accurately estimate abundance from aerial
survey counts of seals basking on ice (i.e., to correct for the proportion of
animals that are in the water while surveys are conducted). Using generalized
linear mixed pseudo-models to properly account for temporal autocorrelation, we
fit models with covariates of interest (e.g., day-of-year, solar hour, age-sex
class, wind speed, barometric pressure, temperature, precipitation) to examine
their ability to explain variation in hourly haul-out probability. We found evidence
for strong diurnal and within-season patterns in haul-out behavior, as well as
strong weather effects (particularly wind and temperature). In general, seals
were more likely to haul out on ice in the middle of the day and when wind speed
was low and temperatures were higher. Haul-out probability increased through
March and April, peaking in May and early June before declining again. The
timing and frequency of haul-out events also varied based on species and age-sex
class. For ribbon and spotted seals, models with year effects were highly
supported, indicating that the timing and magnitude of haul-out behavior varied
among years. However, we did not find evidence that haul-out timing was linked
to annual sea ice extent. Our analysis emphasizes the importance of accounting
for seasonal and temporal variation in haul-out behavior, as well as associated
environmental covariates, when interpreting the number of seals counted
in aerial surveys.

@@

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have ripple effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for
ice-associated seals (ringed (*Pusa hispida*), bearded (*Erignathus barbatus*),
ribbon seals (*Histriophoca fasciata*), and spotted (*Phoca largha*)) that
depend on spring and early summer sea ice (March-June) in the Bering and Chukchi
seas as a platform for important functions, such as pupping, nursing, breeding,
and molting [@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data
and large knowledge gaps complicate predictions about the ultimate effects of
changes in sea ice on the behavior, health, abundance, and distribution of these
seals. To date, indices of seal health sampled during periods of declining sea
ice differ regionally [@crawford2015; @harwood2020]. Knowledge about
evolutionary constraints on the timing of reproductive and molting behavior is
generally lacking, so it is difficult to predict the readiness with which
ice-associated seal species will be able to adapt to future changes (e.g., by
adjusting pupping or molting schedules to earlier dates or different locales).
Additionally, trends in abundance of these species are largely unknown, so it is
difficult to assess the effect, if any, declines in sea ice habitat have had, or
will have, on seal densities.

For ice-associated seals, haul-out behavior is tightly linked with sea ice.
Ribbon seals haul out of the water almost exclusively on sea ice and are mostly
pelagic outside the spring pupping, breeding, and molting season [@boveng2018a].
While spotted and bearded seals rest on coastal features, they
strongly prefer sea ice as a resting platform during the spring and early summer
[@frost2018]. The remoteness of the Bering and Chukchi seas mean direct
scientific observation of seal behavior is impractical. Thus, bio-logging
devices are, thus, especially useful tools for collecting key information on
movement and haul-out behavior for these species.

@@

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of 3 species of ice-associated seals in the Bering and Chukchi seas. We
focused our effort on bearded, ribbon, and spotted seals and did not include
ringed seals in the analysis. Ringed seals are unique in their use of snow lairs
in the spring months, and they warrant a separate study that includes additional
environmental covariates, such as the onset of snow melt and snow depth. Our
goals were threefold. First, we wish to establish baseline estimates for the
chronology of haul-out behavior in the critical spring season for each species
across different age and sex classes. Second, we refined estimates of
haul-out availability corrections for aerial surveys in order to improve
estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not investigate
meteorological variables that have been shown to influence haul-out behavior of
walruses [@reder2003; @udevitz2009]. Third, we assessed the annual variability
in haul-out timing and possible linkage to changes in the extent of seasonal sea
ice between 2005 and 2020. Our work extends the scope of previous haul-out
analyses, includes the influence of meteorological variability, and investigates
the potential impact of a changing icescape on the behavior of these species.

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)`
bio-loggers deployed on bearded, ribbon, and spotted seals were subset to
include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
annual molt, which limits the duration of haul-out data they provide.
Additionally, bio-loggers attached to the head or dorsal region are often dry
while the seal is at the surface and can slightly bias the hourly percent-dry values
reported by the bio-logger. For this study, in cases where both bio-logger types were
deployed, hourly percent dry observations from the flipper tag were preferred.
Sex and age class (non-dependent *young-of-the-year*, sexually immature
*subadults*, and mature *adults*) were determined at the time of deployment by
varying combinations of length, claw annuli [@burns1969; @burns1983;
@mclaren1958a], and pelage characteristics for some species. Field
identification of age class can be inexact, particularly when discerning
subadults from adults. In the case of ribbon seals, pelage differences between
subadults and adults are distinguishable, with subadults often having less
defined ribbons. Bearded seal subadults will often have a spotted pattern in the
pelage that is not seen in adults. Spotted seal pelage is monomorphic and can
not be used to discern age class. Despite these challenges, we feel the age
classifications used in this analysis are useful categories in testing for
age-related effects on haul-out behavior. Seals determined to be less than one
year were classified as young-of-the-year. Subadults are those seals likely
greater than one year of age but not yet sexually mature. Adults are likely
sexually mature and older than four years. For those bio-loggers deployed on
young-of-the-year and transmitting into the next year, the age class was
advanced to subadult on 1 March of the following year. Table
\@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

Tags that fall off due to molt, attachment failure, or seal mortality and remain
on ice or land can still send data to satellites; i.e., the bio-logger will be
on ice or land and dry, therefore, it will record and transmit data suggesting
the seal is hauled out. As such, end times of each deployment were identified by
examining bio-logger locations and dive behavior to determine when bio-loggers
ceased providing data consistent with seal behavior. Data outside of the
deployment start and end times were discarded prior to analysis. After
approximately 9 months, 7 devices deployed on the rear flipper of bearded seals
reported implausible hourly percent-dry data. All data after the first instance
of unrealistic values were censored from this analysis. Figure
\@ref(fig:dataCal) shows a calendar-like distribution of all haul-out data
across the study season for each species. Observations for ribbon and spotted
seals are concentrated in the months of May and June due to the timing of
deployment (April and May) and the timing of molt (May and June). During molt,
seals (and their attached bio-loggers) spend more time out of the water and more
data are transmitted. Molt timing also impacts when many deployments end as any
bio-loggers adhered to the hair will fall off. Bearded seal deployments are more
evenly spread with fewer observations in May and June. The majority of bearded
seal deployments start later in the summer and by May bio-loggers have either
fallen off or the batteries are depleted.

```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4.75, fig.cap = "Distribution of hourly percent-dry biologger data from 1 March to 15 July for each species. Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020. The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, many bio-logger deployments started in April and May."}

ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = col_group, y = row_group, color = n, size = n), stroke = 0, shape = 15) +
  scale_size(range = c(1,3.4), breaks = c(100,500,1000,1500)) +
  scale_color_gradient(low = "#6baed6", high = "#08306b") +
  scale_x_continuous(breaks = c(1,4,8,12,16),
                     labels = c("March", "April", "May", "June", "July")) +
  scale_y_discrete(limits = rev) +
  guides(color = "none",
        size = guide_legend(override.aes = list(color = "#2171b5"),
                            title.position = "bottom", title.hjust = 0.5)
        ) +
  facet_grid(species ~ .) + 
  theme(
        axis.text.y = element_blank(),
        axis.text.x = element_text(hjust = 0),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(x = "", y = "", color = "seal-hours",
       size = "seal hours of bio-logger data") +
  ggtitle("Density of haul-out behavior observations varies from March through July",
          subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")
```

@@

Of key interest in this study is the relationship between haul-out behavior and
weather covariates that vary with space and time. We explored the use of a
continuous-time correlated random walk [@devins.johnson2008] movement model to
predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_=250m, _2_=500m, _1_=1500m, _0_=2500m, _A_=2500m, _B_=2500m) or, when
available, the estimated error radius from the Argos Kalman filter algorithm.
Location estimates from FastLoc GPS were all assigned an error radius of 50m.
Any days where haul-out observations were present but location data were missing
we used the last calculated weighted average daily location, and any days where
the location intersected with land were removed from the data set. We recognized
that bearded and spotted seals haul out on land. However, assessing the
relationship between haul-out behavior and weather covariates and their
availability for aerial surveys on land was outside the scope of this study.
Additionally, any daily locations on land are likely more reflective of
coordinate averaging and measurement error instead of actual use of coastal
features. Figure \@ref(fig:dataMap) shows the spatial distribution weighted
locations with available haul-out behavior data for each species across the
study area.

@@

We modeled data from bio-loggers deployed on bearded, ribbon, and spotted seals
to examine factors affecting haul-out behavior on sea ice in the Bering and
Chukchi seas. Our analysis shows all three species of seal haul out
progressively more through the spring and peak near mid-May to early June before
declining again. This pattern aligns well with what has been previously
documented and confirms our haul-out data are likely capturing population-level
behavioral patterns. Seals preferentially haul out on ice shortly after solar
noon, which allows seals to maximize absorption of solar radiation thought to
facilitate the molting process [@feltz1966]. Interestingly, bearded seals appear
to have two peaks in haul-out activity within a day, one shortly after solar
noon, and one centered near solar midnight. A similar bi-modal pattern has been
seen in ringed seals [@vonduyke2020] and suggests that bearded and ringed seals
may be operating under different constraints than ribbon and spotted seals.
Bearded and ringed seals are distributed across higher latitudes and the
extended daylight hours may allow more flexibility in optimizing resting periods
with foraging. Other factors such as polar bear predation (which is rare for
ribbon and spotted seals in the Bering Sea) may also explain differing haul-out
patterns. Bearded seals also showed a more protracted season of haul-out
behavior compared to ribbon and spotted seals. This aligns with findings from
Thometz et al. [@thometz2021] who observed a mean molting period of 119&pm;2
days and a relatively stable resting metabolic rate during that time. While
ribbon seals were not considered in the same study, spotted and ringed seals
underwent molt periods of 33&pm;4 and 28&pm;6 days and exhibited increased
resting metabolic rates.

Unlike previous analyses of seal haul-out behavior (e.g. [@verhoef2010a],
[@conn2014a]), we also investigated the influence of sex-age class on haul-out
probabilities for all species except bearded seals because of low sample size.
While both ribbon and spotted seals exhibited a unimodal diel haul-out pattern
generally centered around local solar noon, there were key differences across
species, age, and sex that match expectations given what we know of their
ecological behavior. Ribbon seals are not known to form triads and our model
predicts a progression of increased haul-out behavior with females starting
earlier in the season than males. Notably, female ribbon seals spend a large
portion of the day in the water during the pupping period, aligning with the
hypothesis that ribbon seal females continue foraging while nursing. Spotted
seals are known to form triads during the breeding season [@burns2009,
@mcclintock2015] where a female and dependent pup are accompanied on the ice by
a suitor male. The male waits for the female to wean the pup and enter estrus,
and fends off any other potential suitor males. Triad formation results in both
males and females spending a large portion of the day hauled out on ice and a
protracted spring haul-out season for both sexes. Because mating occurs in the
water, females may also be less inclined to interrupt their haul-out with
foraging trips while still nursing the pup. We see this reflected in the
predicted haul-out behavior, with both males and females exhibiting a broad
distribution of time out of the water throughout the solar day and the season.
In the case of both ribbon and spotted seals, subadults are the first to begin
consistent haul-out behavior and follow a typical phocid pattern where subadults
molt first as they do not have any reproductive constraints.

We also investigated the influence of weather on haul-out
probabilities, including wind speed, temperature, barometric pressure,
precipitation, and wind chill. These have been investigated for walruses (e.g.
Udevitz et al. [-@udevitz2009]) and a few select studies of ice-associated seals
[@hamilton2018, @perry2017]. Ribbon seals seemed to be the most influenced by
weather, with wind, temperature, barometric pressure, and precipitation all
being important components of the model. Spotted seals were most affected by
wind and barometric pressures. And, for bearded seals, the model indicated wind
and temperature had the greatest impact. In general, and as might be expected,
seals were more likely to haul-out when daily temperatures were warmer, winds
speeds were lower, barometric pressure was higher, and precipitation was lower.
These weather conditions are likely most beneficial for energetic considerations
and may also provide an optimal situation for predator detection. These results
highlight the importance for haul-out behavior models (and subsequent
calculations of availability during surveys) to incorporate weather covariates.

>>>>

Ice-associated seals rely on sea ice for a variety of activities, including
breeding, molting, pupping, and resting. In the Arctic, many of these activities
occur in spring (April-June) as sea ice begins to melt and retreat northward.
Rapid acceleration of climate change in Arctic ecosystems is therefore of
concern as the quantity and quality of suitable habitat is forecast to decrease.
In this paper, we use data collected from satellite-linked bio-loggers deployed
between 2005 and 2020 to investigate the seasonal timing and environmental
factors affecting haul-out behavior by ice-associated seals. We specifically
focused on bearded (_Erignathus barbatus_), ribbon (_Histriophoca fasciata_),
and spotted seals (_Phoca largha_)) in the Bering and Chukchi seas. Because
ringed seals (_Phoca hispida_) are unique in their use of snow lairs, they were
not included and a separate analysis is warranted. In addition to providing
baseline data on phenology, these data also allow us to quantify 'availability',
which is needed to accurately estimate abundance from aerial survey counts of
these seals basking on ice (i.e., to correct for the proportion of animals that
are in the water while surveys are conducted). Using generalized linear mixed
pseudo-models to properly account for temporal autocorrelation, we fit models
with covariates of interest (e.g., day-of-year, solar hour, age-sex class, wind
speed, barometric pressure, temperature, precipitation) to examine their ability
to explain variation in hourly haul-out probability. We found evidence for
strong diurnal and within-season patterns in haul-out behavior, as well as
strong weather effects (particularly wind and temperature). In general, seals
were more likely to haul out on ice in the middle of the day and when wind speed
was low and temperatures were higher. Haul-out probability increased through
March and April, peaking in May and early June before declining again. The
timing and frequency of haul-out events also varied based on species and age-sex
class. For ribbon and spotted seals, models with year effects were highly
supported, indicating that the timing and magnitude of haul-out behavior varied
among years. However, we did not find evidence that haul-out timing was linked
to annual sea ice extent. Our analysis emphasizes the importance of accounting
for seasonal and temporal variation in haul-out behavior, as well as associated
environmental covariates, when interpreting the number of seals counted in
aerial surveys.

@@

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have ripple effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering and Chukchi seas as a platform for important life history
functions, such as pupping, nursing, breeding, and molting [@boveng2009;
@boveng2013a; @cameron2010; @kelly2010]. Limited data and large knowledge gaps
complicate predictions about the ultimate effects of changes in sea ice on the
behavior, health, abundance, and distribution of these seals. To date, indices
of seal health sampled during periods of declining sea ice differ regionally
[@crawford2015; @harwood2020]. Knowledge about evolutionary constraints on the
timing of reproductive and molting behavior is generally lacking, so it is
difficult to predict the readiness with which ice-associated seal species will
be able to adapt to future changes (e.g., by adjusting pupping or molting
schedules to earlier dates or different locales). This is further complicated by
the spatio-temporal variation in the phenology of these life history events 
within regions and throughout their full ranges. Additionally, trends in
abundance from large scale surveys of these species are unknown, so it is difficult to assess the
effect, if any, declines in sea ice habitat have had, or will have, on seal
densities.

For ice-associated seals, haul-out behavior is tightly linked with sea ice.
Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
on sea ice and are mostly pelagic outside the spring pupping, breeding, and
molting season [@boveng2018a]. While spotted (_Phoca largha_) and bearded
(_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
ice as a resting platform during the spring and early summer [@frost2018].
Ringed seals (_Phoca hispida_) --- not included in this study --- bask on
sea-ice but also within snow lairs built during the spring. The remoteness of
the Bering and Chukchi seas mean direct scientific observation of seal behavior
is impractical. Thus, bio-logging devices are, thus, especially useful tools for
collecting key information on movement and haul-out behavior for these species.

@@

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of 3 species of ice-associated seals in the Bering and Chukchi seas. We
focused our effort on bearded, ribbon, and spotted seals and did not include
ringed seals in the analysis. Ringed seals are unique in their use of snow lairs
in the spring months, and they warrant a separate study that includes additional
environmental covariates, such as the onset of snow melt and snow depth. Our
goals were threefold. First, we wished to establish baseline estimates for the
chronology of haul-out behavior in the critical spring season for each species
across different age and sex classes. Second, we refined estimates of
haul-out availability corrections for aerial surveys in order to improve
estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not investigate
meteorological variables that have been shown to influence haul-out behavior of
walruses [@reder2003; @udevitz2009]. Third, we assessed the annual variability
in haul-out timing and possible linkage to changes in the extent of seasonal sea
ice between 2005 and 2020. Our work extends the scope of previous haul-out
analyses, includes the influence of meteorological variability, and investigates
the potential impact of a changing icescape on the behavior of these species.

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)`
bio-loggers deployed on bearded, ribbon, and spotted seals were subset to
include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
following annual molt, which, depending on deployment date, limits the duration
of haul-out data they provide. Additionally, bio-loggers attached to the head or
dorsal region are often dry while the seal is at the surface and can slightly
bias the hourly percent-dry values reported by the bio-logger. For this study,
in cases where both bio-logger types were deployed, hourly percent dry
observations from the flipper tag were preferred.

Field identification of age class can be inexact, particularly when discerning
subadults from adults. Sex and age class (non-dependent _young-of-the-year_,
sexually immature _subadults_, and mature _adults_) were estimated at the time
of deployment by various combinations of length, claw growth ridges [@burns1969;
@burns1983; @mclaren1958a], and pelage characteristics for some species. In the
case of ribbon seals, subadults often have less distinct ribbons than adults.
Bearded seal subadults will often have a spotted pattern in the pelage that is
not seen in adults. Spotted seal pelage cannot be used to reliably discern
adults from subadults. Despite these challenges, we feel the age classifications
used in this analysis are useful in testing for age-related effects on haul-out
behavior. Seals determined to be less than one year were classified as
young-of-the-year. Subadults are those seals likely greater than one year of age
but not yet sexually mature. Adults are likely sexually mature and older than
four years. For those bio-loggers deployed on young-of-the-year and transmitting
into the next year, the age class was advanced to subadult on 1 March of the
following year. Table \@ref(tab:tblDeploy) provides a summary of these
deployments and data received from them.

@@

Tags that fall off due to molt, attachment failure, or seal mortality and remain
on ice or land can still send data to satellites; i.e., the bio-logger will be
on ice or land and dry, therefore, it will record and transmit data suggesting
the seal is hauled out. As such, end times of each deployment were identified by
examining bio-logger locations and dive behavior to determine when bio-loggers
ceased providing data consistent with seal behavior. Data outside of the
deployment start and end times were discarded prior to analysis. After
approximately 9 months, 7 devices deployed on the rear flipper of bearded seals
reported implausible hourly percent-dry data. All data after the first instance
of unrealistic values were censored from this analysis. Figure
\@ref(fig:dataCal) shows the distribution of all haul-out data
across the study season for each species. Observations for ribbon and spotted
seals are concentrated in the months of May and June due to the timing of
deployment (April and May) and the timing of molt (May and June). During molt,
seals (and their attached bio-loggers) spend more time out of the water and more
data are transmitted. Molt timing also impacts when many deployments end as any
bio-loggers adhered to the hair will fall off. Bearded seal deployments are more
evenly spread with fewer observations in May and June. The majority of bearded
seal deployments start later in the summer and by May bio-loggers have either
fallen off or the batteries are depleted.

```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4.75, fig.cap = "Distribution of hourly percent-dry biologger data from 1 March (upper left) to 15 July (lower right) for each species. Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020. The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, many bio-logger deployments started in April and May."}

ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = col_group, y = row_group, color = n, size = n), stroke = 0, shape = 15) +
  scale_size(range = c(1,3.4), breaks = c(100,500,1000,1500)) +
  scale_color_gradient(low = "#6baed6", high = "#08306b") +
  scale_x_continuous(breaks = c(1,4,8,12,16),
                     labels = c("March", "April", "May", "June", "July"),
                     expand = expansion(mult = c(0.01, 0.025))) +
  guides(color = "none",
        size = guide_legend(override.aes = list(color = "#2171b5"),
                            title.position = "bottom", title.hjust = 0.5)
        ) +
  facet_grid(species ~ .) + 
  theme(
        axis.text.y = element_blank(),
        axis.text.x = element_text(hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(color = "seal-hours",
       size = "seal hours of bio-logger data") +
  ggtitle("Density of haul-out behavior observations varies from March through July",
          subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")
```

@@

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with space and time. We explored the use of a
continuous-time correlated random walk [@devins.johnson2008] movement model to
predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_=250m, _2_=500m, _1_=1500m, _0_=2500m, _A_=2500m, _B_=2500m) or, when
available, the estimated error radius from the Argos Kalman filter algorithm.
Location estimates from FastLoc GPS were all assigned an error radius of 50m.
Any days where haul-out observations were present but location data were missing
we used the last calculated weighted average daily location, and any days where
the location intersected with land were removed from the data set. We recognized
that bearded and spotted seals haul out on land. However, assessing the
relationship between haul-out behavior and weather covariates and their
availability for aerial surveys on land was outside the scope of this study.
Additionally, any daily locations on land are likely more reflective of
coordinate averaging and measurement error instead of actual use of coastal
features. Figure \@ref(fig:dataMap) shows the spatial distribution weighted
locations with available haul-out behavior data for each species across the
study area.

@@

We modeled data from bio-loggers deployed on bearded, ribbon, and spotted seals
to examine factors affecting haul-out behavior on sea ice in the Bering and
Chukchi seas. Our analysis shows all three species of seal haul out
progressively more through the spring and peak near mid-May to early June before
declining again. This pattern aligns well with what has been previously
documented and confirms our haul-out data are likely capturing population-level
behavioral patterns. Seals preferentially haul out on ice shortly after solar
noon, which allows seals to maximize absorption of solar radiation thought to
facilitate the molting process [@feltz1966]. Interestingly, bearded seals appear
to have two peaks in haul-out activity within a day, one shortly after solar
noon, and one centered near solar midnight. This, of course, could be an
artifact of our limited sample size for bearded seal deployments across all age
classes. However, a similar bi-modal pattern has been seen in ringed seals
[@vonduyke2020] and suggests that bearded and ringed seals may be operating
under different constraints than ribbon and spotted seals. Bearded and ringed
seals are distributed across higher latitudes and the extended daylight hours
may allow more flexibility in optimizing resting periods with foraging. Other
factors such as predation by polar bears (which is rare for ribbon and spotted
seals in the Bering Sea) may also explain differing haul-out patterns. Bearded
seals also showed a more protracted season of haul-out behavior compared to
ribbon and spotted seals. This aligns with findings from Thometz et al.
[@thometz2021] who observed a mean molting period of 119&pm;2 days and a
relatively stable resting metabolic rate during that time. While ribbon seals
were not considered in the same study, spotted and ringed seals underwent molt
periods of 33&pm;4 and 28&pm;6 days and exhibited increased resting metabolic
rates.

Unlike previous analyses of seal haul-out behavior (e.g. [@verhoef2010a],
[@conn2014a]), we also investigated the influence of sex-age class on haul-out
probabilities for all species except bearded seals because of low sample size.
While both ribbon and spotted seals exhibited a unimodal diel haul-out pattern
generally centered around local solar noon, there were key differences across
species, age, and sex that match expectations given what we know of their
ecological behavior. Ribbon seals are not known to form triads and our model
predicts a progression of increased haul-out behavior with females starting
earlier in the season than males. Notably, female ribbon seals spend a large
portion of the day in the water during the pupping period, aligning with the
hypothesis that ribbon seal females continue foraging while nursing. Spotted
seals are known to form triads during the breeding season [@burns1973] where a
female and dependent pup are accompanied on the ice by a suitor male. The male
waits for the female to wean the pup and enter estrus, and fends off any other
potential suitor males. Triad formation results in both males and females
spending a large portion of the day hauled out on ice and a protracted spring
haul-out season for both sexes. Because mating occurs in the water, females may
also be less inclined to interrupt their haul-out with foraging trips while
still nursing the pup. We see this reflected in the predicted haul-out behavior,
with both males and females exhibiting a broad distribution of time out of the
water throughout the solar day and the season. In the case of both ribbon and
spotted seals, subadults are the first to begin consistent haul-out behavior and
follow a typical phocid pattern where subadults molt first as they do not have
any reproductive constraints.

We also investigated the influence of weather on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind
chill. These have been investigated for walruses (e.g. Udevitz et al.
[-@udevitz2009]) and a few select studies of ice-associated seals
[@hamilton2018, @perry2017]. Ribbon seals seemed to be the most influenced by
weather, with wind, temperature, barometric pressure, and precipitation all
being important components of the model. Spotted seals were most affected by
wind and barometric pressures. And, for bearded seals, the model indicated wind
and temperature had the greatest impact. In general, and as might be expected,
seals were more likely to haul-out when daily temperatures were warmer, winds
speeds were lower, barometric pressure was higher, and precipitation was lower.
These weather conditions are general indicators of increased solar radiation
which provides energetic benefits. Low winds and precipitation may also
provide an optimal situation for predator detection. Our results highlight the
importance for haul-out behavior models (and subsequent calculations of
availability during surveys) to incorporate weather covariates.
